---
title: "Otsu-guided segmentation"
format:
  html:
    toc: true
    code-copy: true
engine: knitr
editor: source
execute:
  echo: true
  output: true
---

```{=html}
<style>
.justified p {
  text-align: justify;
}
</style>
```

## Otsu-guided segmentation and rule-based filtering

::: {.justified}

This is the second stage of the workflow. It finds candidate burned
patches with rules only: no training data, no model.

The change-index mosaic from the previous chapter is split into
burned / unburned by an Otsu threshold. The threshold is not global:
it is computed separately for every combination of land-cover class
and ecoregion, so a pine forest in the north and a shrubland in the
south are cut at different severities.

Pixels above a strict "seed" threshold start a candidate patch; the
patch then grows into neighbouring pixels above a looser threshold.
Grown patches pass through a set of filters (size, overlap with the
previous year's fires, spectral support), and each one ends up
labelled `keep`, `review` or `drop`.

The output that matters is `internal_decisions.gpkg`: one polygon per
candidate, with its label. This file is the training input for the
next stage. Everything else this stage writes is diagnostics.

In the API these functions are called `deterministic`, because this
stage follows explicit rules rather than fitting a classifier. The
Otsu thresholds themselves are not fixed: they are estimated from the
input data.

:::

## 1. Load OtsuFire

```{r}
library(OtsuFire)
```

This stage delegates patch growing to the
[WhiteboxTools](https://www.whiteboxgeo.com/) engine, via the
`whitebox` R package. The first time you use it, download the
underlying binaries once with `whitebox::install_whitebox()`.

## 2. Locate your input data

Segmentation needs more inputs than the mosaic step:

- `change_index`: the mosaic built in the previous chapter.
- `vegetation_map`: a land-cover raster (e.g. CORINE), used to vary
  the Otsu threshold by class.
- `burnable_mask`: a binary raster on the same grid, marking pixels
  that can burn at all.
- `ecoregion_shapefile_path` / `peninsula_shapefile_path`: the spatial
  units used to compute per-unit thresholds, and the overall study
  area.
- `hotspots` and `previous_year_burned` are optional: active-fire
  detections and last year's burned area. Set either to `NULL` if you
  do not have it for your target year.

Replace the paths below with the locations on your computer.

```{r}
change_index      <- "data/mosaics/change_index_2013.tif"
vegetation_map     <- "data/land_cover/vegetation_map.tif"
burnable_mask      <- "data/land_cover/burnable_mask.tif"
ecoregion_shp      <- "data/boundaries/ecoregions.shp"
peninsula_shp      <- "data/boundaries/study_area.shp"

previous_year_burned <- "data/fires/burned_2012.shp"  # NULL if unavailable
hotspots             <- "data/hotspots/hotspots_2013.geojson"  # NULL if unavailable
```

```{r}
#| include: false
if (file.exists("_local_paths.R")) source("_local_paths.R")
```

## 3. Choose the detection parameters

Detection is controlled by three parameter lists, instead of a closed
set of named presets.

**`detect_params`** decides where a patch starts and how far it grows:

- `seed_threshold`: the severity a pixel needs to start a patch.
  Higher means fewer, more certain patches.
- `growth_delta`: how far below the seed threshold a neighbouring
  pixel may be and still be absorbed into the patch.
- `minimum_growth_threshold`: a hard floor that growth never crosses,
  whatever `growth_delta` allows.
- `minimum_seed_pixels`: the minimum number of seed pixels a patch
  needs to exist. This is the main noise filter.

Each of these can also be supplied as a `*_by_vegetation` vector,
overriding the global value per land-cover class. Use this when the
same severity means different things in different vegetation: a
moderate change in shrubland can be a real fire, while the same value
in a dense forest usually is not.

```{r}
detect_params <- list(
  seed_threshold           = 300,
  growth_delta             = 100,
  minimum_growth_threshold = 200,
  minimum_seed_pixels      = 20L
)
```

**`refine_params`** cleans up the resulting geometry: a working buffer
around the study area, a minimum area below which slivers are
dropped, and whether overlapping patches are merged.

```{r}
refine_params <- list(
  aoi_buffer_m             = 5000,
  minimum_detected_area_m2 = 10,
  merge_overlaps           = TRUE
)
```

**`scoring_params`** controls the filters that assign the final
`keep` / `review` / `drop` label: how a patch is penalised for sitting
on the previous year's scar, and the minimum area it must retain after
that penalty.

```{r}
scoring_params <- list(
  previous_fire_exclusion_buffer_m = 90,
  minimum_remaining_area_m2        = 20000
)
```

See `?build_burned_mapping_config` for the complete list of settings
in each block, and the book's reference chapter for guidance on
tuning them for your own study area.

## 4. Build the configuration

As in the mosaic stage, building the configuration validates your
inputs immediately; it does not run anything yet.

```{r}
det_cfg <- build_burned_mapping_config(
  change_index          = change_index,
  vegetation_map         = vegetation_map,
  burnable_mask          = burnable_mask,
  hotspots               = hotspots,
  previous_year_burned   = previous_year_burned,
  target_year            = 2013L,
  output_dir             = "results",
  run_name               = "demo_run",
  detect_params  = detect_params,
  refine_params  = refine_params,
  scoring_params = scoring_params,
  options = list(
    ecoregion_shapefile_path = ecoregion_shp,
    peninsula_shapefile_path = peninsula_shp
  )
)
```

## 5. Run detection and scoring

One call runs detection and scoring together. Expect this to take
tens of minutes for a full study area and year, depending on its size
and the number of candidate patches.

```{r}
#| eval: false

det <- run_deterministic_pipeline(
  det_cfg,
  write_outputs  = TRUE,
  overwrite      = FALSE,
  run_validation = FALSE  # the next chapter handles validation
)
```

The paths it writes:

```{r}
#| eval: false

det$result_paths$grown_patches
det$result_paths$refined_patches
det$result_paths$internal_decisions  # <- the input to the next stage
det$result_paths$timing_csv
```

## 6. Check the result

```{r}
#| eval: false

dec <- sf::st_read(det$result_paths$internal_decisions, quiet = TRUE)

nrow(dec)
table(dec$class_final, useNA = "ifany")  # keep / review / drop counts

# A useful sanity check: total kept area, in hectares.
sum(as.numeric(sf::st_area(dec[dec$class_final == "keep", ])) / 10000)
```

If `keep` is a handful of polygons, the thresholds are too strict for
this year and study area. If it is tens of thousands, they are too
loose. Adjust `seed_threshold` and `minimum_seed_pixels` first: they
move the result the most.

## 7. Step by step

`run_deterministic_pipeline()` is a wrapper over two functions. Call
them directly when you want to inspect the candidates before they are
scored.

```{r}
#| eval: false

patches <- detect_burned_patches(det_cfg, write_outputs = TRUE)
patches$grown_patches_path

scored <- score_burned_patches(patches, det_cfg, write_outputs = TRUE)
scored$internal_decisions_path
```

## 8. Optional: separating chained fires

Two real fires can be joined by a thin neck of burned pixels and come
out as a single, oddly shaped patch. `apply_chain_cleaning()` cuts
necks narrower than a chosen width and splits such a patch back into
its parts. It is opt-in, and it leaves well-shaped patches untouched.

```{r}
#| eval: false

chained <- apply_chain_cleaning(
  refine_path        = det$result_paths$refined_patches,
  year               = 2013L,
  dechain_distance_m = 180,
  pixel_size_m       = 90
)
```

Score the de-chained patches instead of the original ones if you use
this step.

Keep the `internal_decisions.gpkg` path: it is the input for the
supervised stage, in the next chapter. Its `keep` / `review` / `drop`
labels are rules, not ground truth, and the next stage learns from
them.
